The Error Estimate of Force Calculation in the Inhomogeneous Molecular Systems

### 2012-06-03 longRange.hetero/alltext/writeDown/main.tex
mv dirs to alltext

\section{Theory}
\subsection{Ewald summation}
The Ewald method divides the electrostatic interaction in to three
parts: the direct part, the reciprocal part and the correction
part. that is
\begin{equation}
E = E_{\textrm{dir}} + E_{\textrm{rec}} + E_{\textrm{corr}},
\end{equation}
where 
\begin {align}
E_{\textrm{dir}} & = \frac12 \sum^{\ast}_{\v n}
\sum_{i,j = 1}^{N} \frac{q_iq_j \textrm{erfc}(\beta \vert\v{r}_{ij} + \v{n}\vert)}
{\vert\v{r}_{ij} + \v{n}\vert} \\
E_{\textrm{rec}} & = \frac1{2\pi V} \sum_{\v m \neq 0}
\frac{\exp(-\pi^2\v m^2 / \beta^2)}{\v m^2} S(\v m) S(-\v m) \\
 E_{\textrm{corr}}& = -\frac\beta{\sqrt \pi} \sum_{i=1}^N q_i^2
\end {align}
The distance between the two particles is denoted by $\v r_{ij} = \v
r_i - \v r_j$, where $\v r_i$ and $\v r_j$ are positions of particle
$i$ and $j$, respectively.  The lattice in the real space is denoted
by $\v n = n_1 \v a_1 + n_2 \v a_2 + n_3 \v a_3$, where $\v a_\alpha,
\ \alpha=1,2,3$ are box vectors. the structure factor $S(\v m)$ is
defined by
\begin{equation}\label{sm1}
S(\v m) = \sum_{j=1}^N q_j \exp (2 \pi i \v m \cdot \v r_j),
\end{equation}
where $\v m = m_1 \v a_1^\ast + m_2 \v a_2^\ast + m_3 \v a_3^\ast$ is
the reciprocal space lattice vectors. $\v a_\alpha^\ast,\ \alpha = 1,
2, 3$ are conjugate reciprocal vectors of $\v a_\alpha$, defined by
$\v a_\alpha \cdot \v a_\beta^\ast = \delta_{\alpha\beta}$. and $V =
\v a_1 \cdot \v a_2 \times \v a_3$ is the volume of the box.

Both the direct and the reciprocal parts are calculated by cut-off in
real and reciprocal space, respectively. For the reciprocal part the
cut-offed energy reads
\begin{align}
  E_{\textrm{rec}} & =
  \frac1{2\pi V}
  \sum_{
    \begin{subarray}{c}
      \vert m_\alpha\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \frac{\exp(-\pi^2\v m^2 / \beta^2)}{\v m^2} S(\v m) S(-\v m)   
\end{align}
where $K_\alpha$ is the number of Fourier modes used on direction
$\alpha$. For convenience, we denote
\begin{align}
  f(\v m) = \frac{\exp(-\pi^2 \v m^2 / \beta^2)}{2\pi V \v m^2}
\end{align}
The truncated reciprocal force acting on particle $i$ is
\begin{align}
  \v F(\v r_i) = - 
  q_i 
  \sum_{
    \begin{subarray}{c}
      \vert m_\alpha\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \v g(\v m) \,
  S(-\v m)\,
  e^{2\pi i\v m\cdot\v r_i}
\end{align}
where
\begin{align}
  \v g(\v m) = -4 \pi \v m i f(\v m)
\end{align}

\subsection{Error estimate for the reciprocal Ewald force}

In this paper, we consider the force error, bacause the applicaions
are based on the molecular dynamics simulation.  Suppose the system is
composed by $N$ charged particles located at $\v r_1, \v r_2, \cdots,
\v r_N$ with charges $q_1, q_2, \cdots, q_N$, respectively.  We study
the force error of an arbitrary charge $i$ by the difference between
the exact force $\v F(\v r_i)$ and calculated force $\tilde{\v F}(\v
r_i)$ (with error introduced by cut-offs) exerting on the charge $i$.
The difference between the exact and calculated is called the
\emph{error force} \cite{short}, and is denoted by $\Delta \v F(\v r_i)
= \v F(\v r_i) - \tilde{\v F}(\v r_i)$.  The error of the force
computation is defined by the \emph{root mean square (RMS)}
\emph{error}, which is the square root of the second moment of the
error force: $\mathcal E(\v r_i) = \sqrt{\langle\vert\Delta\v F(\v
  r_i)\vert^2\rangle}$.  The notation $\langle\cdot\rangle$ means
ensemble average.

Ref.~\cite{wang2010optimizing} pointed out that it
is reasonable and convenient to estimate the error for the direct and
reciprocal force part seperately. 
The direct part is a short-range
interaction in real space, the error estimate of which
is well established in \cite{short}.
The mean of the error force is
\begin{align}\label{eqn:tmp10}
  \langle\Delta\v F_{\textrm{dir}}(\v r_i)\rangle
  =q_i\, [\v K^c_{\textrm{dir}}\ast\rho_q](\v r_i).
\end{align}
Herein, the ``$\ast$'' notation means convolution.  The direct
convolutional kernel of the error force $\v K^c_{\textrm{dir}}(\v r)$
is defined by
\begin{align}
  \v K^c_{\textrm{dir}}(\v r) =
  \left\{
    \begin{aligned}
      &\,0, &\quad & \vert\v r\vert \leq r_c;\\
      &,& & \vert\v r\vert > r_c,
      % \frac{\textrm{erfc}(\beta\vert\v r\vert)}{\vert\v r\vert}
    \end{aligned}
  \right.
\end{align}
where $r_c$ is the cut-off radius in the real space.  The first order
charge distribution function $\rho_q$ is defined by:
\begin{align}
  \rho_q(\v r) = 
  \bigg\langle
  \sum_{j = 1}^N
  \,q_j\,\delta(\v r - \v r_j)
  \bigg\rangle.
\end{align}
With the help of fast Fourier transform (FFT), the convolution in
\eqref{eqn:tmp10} can be calculated with the computational cost of
$\mathcal O(N\log N)$.

And the RMS error is estimated by
\begin{align}\nonumber
  \langle\vert\Delta\v F_{\textrm{dir}}(\v r_i)\vert^2\rangle
  = &\,
  q_i^2\,[(\v K^c_{\textrm{dir}})^2\ast\rho_{q^2}] (\v r_i) + 
  q_i^2\,[\v K^c_{\textrm{dir}}\ast\rho_{q}]^2 (\v r_i) + \\
  &
  q_i^2\,\int_{\mathbb R^3\times\mathbb R^3}
  \v K^c_{\textrm{dir}}(\v r_i-\v r')\cdot
  \v K^c_{\textrm{dir}}(\v r_i-\v r'')\,
  C_{q^2}(\v r', \v r'')\,
  \d d\v r'\d d\v r''.
\end{align}
The three terms on the right hand side are homogeneity error,
inhomogeneity error and correlation error. The inhomogeneity error is
nothing but the square of the mean error force \eqref{eqn:tmp10}. The
second order charge distribution $\rho_{q^2}$ is
\begin{align}
  \rho_{q^2}(\v r) = 
  \bigg\langle
  \sum_{j = 1}^N
  \,q^2_j\,\delta(\v r - \v r_j)
  \bigg\rangle.
\end{align}
The charge density correlation function is defined by
\begin{align}
  C_{q^2}(\v r', \v r'')
  =
  \bigg\langle
  \sum_{i\neq j}q_iq_j\delta(\v r'-\v r_i)\delta(\v r''-\v r_j)
  \bigg\rangle
  - \rho_q(\v r')\rho_q(\v r'').
\end{align}
In most situations, neglecting the correlation error will give
satisfactory estimate of the error~\cite{short}. Therefore, 
\begin{align}
  \langle\vert\Delta\v F_{\textrm{dir}}(\v r_i)\vert^2\rangle
  = &\,
  q_i^2\,[(\v K^c_{\textrm{dir}})^2\ast\rho_{q^2}] (\v r_i) + 
  q_i^2\,[\v K^c_{\textrm{dir}}\ast\rho_{q}]^2 (\v r_i).
\end{align}

The reciprocal error force, namely the difference between the
the exact and truncated reciprocal force, is
\begin{align}
  \Delta \v F_{\textrm{rec}}(\v r_i)
  = & - 
  q_i
  \sum_{\vert m_\alpha\vert \geq K_\alpha/2}
  \v g(\v m) \,
  S(-\v m)\,
  e^{2\pi i\v m\cdot\v r_i} 
\end{align}
It is proved that the mean reciprocal error force is
\begin{align}\label{eqn:tmp18}
  \big\langle
  \Delta \v F_{\textrm{rec}}(\v r_i)
  \big\rangle
  &= 
  -q_i\,[\v K^c_{\textrm{rec}}\ast\rho_q\,] (\v r_i),
\end{align}
where the reciprocal convolutional kernel of the error force $\v
K^c_{\textrm{rec}}$ is defined by:
\begin{align}
  \v K^c_{\textrm{rec}}(\v r) =
  \sum_{
      \vert m_\alpha\vert \geq K_\alpha/2}
  \v g(\v m) \,
  e^{2\pi i\v m\cdot\v r}.
\end{align}
Similarly, with the assumption of uncorrelation, it is shown that
\begin{align}
  \langle\vert\Delta\v F_{\textrm{rec}}(\v r_i)\vert^2\rangle
  = &\,
  q_i^2\,[(\v K^c_{\textrm{rec}})^2\ast\rho_{q^2}] (\v r_i) + 
  q_i^2\,[\v K^c_{\textrm{rec}}\ast\rho_{q}]^2 (\v r_i).
\end{align}

@@

\section{SPME}

The principle idea of the SPME method is to approximate the term
$e^{2\pi i\v m\cdot\v r}$ in the structure factor by the B-spline
interpolation. Denote the error introduced by this approximation by
$A(\v m, \v r)\,e^{2\pi i\v m\cdot \v r}$, we are fortunate to have
the analytical expression of $A(\v m, \v r)$:
\begin{align}
  A(\v m, \v r)
  =
  \sum_{\alpha=1}^3\sum_{l\neq 0}
  \frac
  {\big(\dfrac{2\pi m_\alpha}{K_\alpha} + 2\pi l\big)^{-n}}
  {\sum\limits_{l=-\infty}^{+\infty}
    \big(\dfrac{2\pi m_\alpha}{K_\alpha} + 2\pi l\big)^{-n}}
  \bigg(
  e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r} - 1
  \bigg)
\end{align}

There are two possibility of calculating the reciprocal force from the
SPME energy. The first one is to differentiate the reciprocal term of
the original Ewald summation and then approximated the analytical
force by the B-spline interplation. This way is called
\emph{ik--differentiation}. The other way is to negative gradient on
the SPME approximated energy, as proposed in the original SPME paper
\cite{essmann1995spm}. The later is called \emph{analytical
differentiation}.

\subsection{Error estimate of the ik-differentiation}
The error force is 
\begin{align}\nonumber
  \Delta\v F^{\textrm{ik}}(\v r_i)
  =\;&
  q_i
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \v g(\v m)\,
  A(\v m, \v r_i)\,
  e^{2\pi i\v m\cdot \v r_i}
  \sum_{j\neq i}q_j\,e^{-2\pi i\v m\cdot\v r_j} \\\nonumber
  & +\,
  q_i
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \v g(\v m)\,
  e^{2\pi i\v m\cdot \v r_i}
  \sum_{j\neq i}q_j\,A(\v m, -\v r_j)\,e^{-2\pi i\v m\cdot\v r_j}\\\nonumber
  & +\,
  q_i
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \v g(\v m)\,
  \big[
  A(\v m, \v r_i) +
  A(\v m, -\v r_i)
  \big]
\end{align}
The last ``self-interacting term'' is zero because of the symmetricity
of $\v g(\v m)$ and $A(\v m, \v r)$. 
By denoting
\begin{align}
  \v G_{\alpha,l}(\v r) =
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \v g(\v m)\,
  \frac
  {\big(\dfrac{2\pi m_\alpha}{K_\alpha} + 2\pi l\big)^{-n}}
  {\sum\limits_{l=-\infty}^{+\infty}
    \big(\dfrac{2\pi m_\alpha}{K_\alpha} + 2\pi l\big)^{-n}}
  e^{2\pi i\v m\cdot\v r}
\end{align}
And if the high wave number terms are neglected, it is proved that the
mean error force is
\begin{align}
  \big\langle
  \Delta\v F^{\textrm{ik}}(\v r_i)
  \big\rangle
  & \approx
  - 2 q_i
  \bigg[
  \big(
  \sum_{\alpha}\sum_{l\neq 0}\v G_{\alpha,l}
  \big)
  \ast\rho_q\,
  \bigg] (\v r_i)
\end{align}
Also neglect the high wave number terms, the mean squared error force is 
\begin{align}\nonumber
  \big\langle
  \vert\Delta\v F^{\textrm{ik}}(\v r_i)\vert^2
  \big\rangle
  \approx&\, 
  2q_i^2
  \bigg[\,
  \big(
  \sum_{\alpha} \sum_{l\neq 0}
  \vert \v G_{\alpha,l}\vert^2
  \big)
  \ast \rho_{q^2}
  \,\bigg] (\v r_i)
  +
  4q_i^2\,
  \bigg[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)^2
  \ast \rho_{q^2}\,
  \bigg] (\v r_i) \\
  &\,+
  \big\langle
  \Delta\v F^{\textrm{ik}}(\v r_i)
  \big\rangle  ^2
\end{align}

\subsection{Analytical differetiation}

The error force is 

@@

  \Delta\v F^{\textrm{ana}}(\v r_i)

@@

  \Delta\v F^{\textrm{ik}}(\v r_i)\\\nonumber

@@

  \langle \Delta\v F^{\textrm{ana}}(\v r_i)\rangle

@@

  \langle \Delta\v F^{\textrm{ik}}(\v r_i)\rangle\\\nonumber

@@

  \langle \Delta\v F^{\textrm{ik}}(\v r_i)\rangle,

@@

  \vert \Delta\v F^{\textrm{ana}}(\v r_i)\vert^2

@@

  \big\langle \Delta\v F^{\textrm{ana}}(\v r_i)\big\rangle ^2

@@

  \big\langle \Delta\v F^{\textrm{ik}}(\v r_i)\big\rangle ^2 \\\nonumber

@@

  \big\langle \Delta\v F^{\textrm{ik}}(\v r_i)\big\rangle ^2\\

@@

\section{Numerical tests}

>>>>

\section{Theory}
\subsection{Ewald summation}
The Ewald method divides the electrostatic interaction in to three
parts: the direct part, the reciprocal part and the correction
part. that is
\begin{equation}
E = E_{\textrm{dir}} + E_{\textrm{rec}} + E_{\textrm{corr}},
\end{equation}
where 
\begin {align}\label{eqn:tmp2}
E_{\textrm{dir}} & = \frac12 \sum^{\ast}_{\v n}
\sum_{i,j = 1}^{N} \frac{q_iq_j \textrm{erfc}(\beta \vert\v{r}_{ij} + \v{n}\vert)}
{\vert\v{r}_{ij} + \v{n}\vert} \\\label{eqn:tmp3}
E_{\textrm{rec}} & = \frac1{2\pi V} \sum_{\v m \neq 0}
\frac{\exp(-\pi^2\v m^2 / \beta^2)}{\v m^2} S(\v m) S(-\v m) \\\label{eqn:tmp4}
 E_{\textrm{corr}}& = -\frac\beta{\sqrt \pi} \sum_{i=1}^N q_i^2
\end {align}
The distance between the two particles is denoted by $\v r_{ij} = \v
r_i - \v r_j$, where $\v r_i$ and $\v r_j$ are positions of particle
$i$ and $j$, respectively.  The lattice in the real space is denoted
by $\v n = n_1 \v a_1 + n_2 \v a_2 + n_3 \v a_3$, where $\v a_\alpha,
\ \alpha=1,2,3$ are box vectors. the structure factor $S(\v m)$ is
defined by
\begin{equation}\label{sm1}
S(\v m) = \sum_{j=1}^N q_j \exp (2 \pi i \v m \cdot \v r_j),
\end{equation}
where $\v m = m_1 \v a_1^\ast + m_2 \v a_2^\ast + m_3 \v a_3^\ast$ is
the reciprocal space lattice vectors. $\v a_\alpha^\ast,\ \alpha = 1,
2, 3$ are conjugate reciprocal vectors of $\v a_\alpha$, defined by
$\v a_\alpha \cdot \v a_\beta^\ast = \delta_{\alpha\beta}$. and $V =
\v a_1 \cdot \v a_2 \times \v a_3$ is the volume of the box.

The complementary error function $\textrm{erfc}(r)$ in
Eqn.~\eqref{eqn:tmp2} decays exponentially as $r$ goes to infinity.
Therefore, it is a short-range interaction in the real space, which
can be calculated by the standard neighbor list
method~\cite{frenkel02b} at a cost of $\mathcal O(N)$.  The reciprocal
part also decays exponentially as the maganitude of the Fourier mode
$\vert\v m\vert$ increases. Therefore, in practice, the infinit
summation in Eqn.~\eqref{eqn:tmp3} is truncated, and only a finite
summation is calculated:
% Both the direct and the reciprocal parts are calculated by cut-off in
% real and reciprocal space, respectively.
% For the reciprocal part the cut-offed energy reads
\begin{align}\label{eqn:tmp6}
  E^{\textrm{tr}}_{\textrm{rec}} & =
  % \frac1{2\pi V}
  \sum_{
    \begin{subarray}{c}
      \vert m_\alpha\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  % \frac{\exp(-\pi^2\v m^2 / \beta^2)}{\v m^2} S(\v m) S(-\v m)   
  f(\v m)\, S(\v m) S(-\v m),
\end{align}
where
\begin{align}
  f(\v m) = \frac{\exp(-\pi^2 \v m^2 / \beta^2)}{2\pi V \v m^2},
\end{align}
and $K_\alpha$ is the number of Fourier modes used on direction
$\alpha$.  The truncated reciprocal force acting on particle $i$ is
\begin{align}\label{eqn:tmp8}
  \v F^{\textrm{tr}}_{\textrm{rec}}(\v r_i) = - 
  q_i 
  \sum_{
    \begin{subarray}{c}
      \vert m_\alpha\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \v g(\v m) \,
  S(-\v m)\,
  e^{2\pi i\v m\cdot\v r_i},
\end{align}
where
\begin{align}
  \v g(\v m) = -4 \pi \v m i f(\v m).
\end{align}

\subsection{Error estimate for the Ewald summation}

In this paper, we consider the force error, bacause the applicaions
are based on the molecular dynamics simulation.  Suppose the system is
composed by $N$ charged particles located at $\v r_1, \v r_2, \cdots,
\v r_N$ with charges $q_1, q_2, \cdots, q_N$, respectively.  We study
the force error of an arbitrary charge $i$ by calculating the
difference between the exact force and calculated force exerting on
the charge $i$, which is called the \emph{error force}
\cite{wang2012} and denoted by $\Delta \v F(\v r_i)$.
% and is denoted by
% $\Delta \v F(\v r_i) = \v F(\v r_i) - \tilde{\v F}(\v r_i)$.
The maganitude of the force computation error 
% error of the force computation
is defined by the \emph{root mean square (RMS)} \emph{error}, which is
the square root of the second moment of the error force: $\mathcal
E(\v r_i) = \sqrt{\langle\vert\Delta\v F(\v r_i)\vert^2\rangle}$.  The
notation $\langle\cdot\rangle$ means ensemble average.
Ref.~\cite{wang2010optimizing} pointed out that it is reasonable and
convenient to estimate the error for the direct and reciprocal part
seperately, namely $\Delta\v F(\v r_i) = \Delta\v F_{\textrm{dir}}(\v
r_i) + \Delta\v F^{\textrm{tr}}_{\textrm{rec}}(\v r_i)$.  The
$\Delta\v F_{\textrm{dir}}$ and $ \Delta\v
F^{\textrm{tr}}_{\textrm{rec}}$ denote the direct error force and
reciprocal error force, respectively.  The superscript ``tr'' denotes
the reciprocal force is calculated by truncating the infinite Ewald
summation. Other superscripts are introduced later in the paper to
denote the error force of the fast algoritms
that treat the Ewald summation more elaborately.

The direct part is a short-range
interaction in real space, the error estimate of which
is well established in \cite{wang2012}.
The mean of the error force is
\begin{align}\label{eqn:tmp10}
  \langle\Delta\v F_{\textrm{dir}}(\v r_i)\rangle
  =q_i\, [\v K^c_{\textrm{dir}}\ast\rho_q](\v r_i).
\end{align}
Herein, the ``$\ast$'' notation means convolution.  The direct
convolutional kernel of the error force $\v K^c_{\textrm{dir}}(\v r)$
is defined by
\begin{align}
  \v K^c_{\textrm{dir}}(\v r) =
  \left\{
    \begin{aligned}
      &\,0, &\quad & r \leq r_c;\\
      &
      \Big[
      \frac{2\beta}{\sqrt\pi} e^{-\beta^2r^2} + \frac{\textrm{erfc}(\beta r)}{r}
      \Big]\frac{\v r}{r^2}
      ,& & r > r_c,
      % \frac{\textrm{erfc}(\beta\vert\v r\vert)}{\vert\v r\vert}
    \end{aligned}
  \right.
\end{align}
where $r = \vert\v r\vert$, and  $r_c$ is the cut-off radius in the real space.  The first order
charge distribution function $\rho_q$ is defined by:
\begin{align}
  \rho_q(\v r) = 
  \bigg\langle
  \sum_{j = 1}^N
  \,q_j\,\delta(\v r - \v r_j)
  \bigg\rangle.
\end{align}
With the help of fast Fourier transform (FFT), the convolution in
\eqref{eqn:tmp10} can be calculated with the computational cost of
$\mathcal O(N\log N)$.

And the RMS error is estimated by
\begin{align}\nonumber
  \langle\vert\Delta\v F_{\textrm{dir}}(\v r_i)\vert^2\rangle
  = &\,
  q_i^2\,[(\v K^c_{\textrm{dir}})^2\ast\rho_{q^2}] (\v r_i) + 
  q_i^2\,[\v K^c_{\textrm{dir}}\ast\rho_{q}]^2 (\v r_i) + \\
  &
  q_i^2\,\int_{\mathbb R^3\times\mathbb R^3}
  \v K^c_{\textrm{dir}}(\v r_i-\v r')\cdot
  \v K^c_{\textrm{dir}}(\v r_i-\v r'')\,
  C_{q^2}(\v r', \v r'')\,
  \d d\v r'\d d\v r''.
\end{align}
The three terms on the right hand side are homogeneity error,
inhomogeneity error and correlation error. The inhomogeneity error is
nothing but the square of the mean error force \eqref{eqn:tmp10}. The
second order charge distribution $\rho_{q^2}$ is
\begin{align}
  \rho_{q^2}(\v r) = 
  \bigg\langle
  \sum_{j = 1}^N
  \,q^2_j\,\delta(\v r - \v r_j)
  \bigg\rangle.
\end{align}
The charge density correlation function is defined by
\begin{align}
  C_{q^2}(\v r', \v r'')
  =
  \bigg\langle
  \sum_{i\neq j}q_iq_j\delta(\v r'-\v r_i)\delta(\v r''-\v r_j)
  \bigg\rangle
  - \rho_q(\v r')\rho_q(\v r'').
\end{align}
In most situations, neglecting the correlation error will give
satisfactory estimate of the error~\cite{wang2012}. Therefore, 
\begin{align}
  \langle\vert\Delta\v F_{\textrm{dir}}(\v r_i)\vert^2\rangle
  = &\,
  q_i^2\,[(\v K^c_{\textrm{dir}})^2\ast\rho_{q^2}] (\v r_i) + 
  q_i^2\,[\v K^c_{\textrm{dir}}\ast\rho_{q}]^2 (\v r_i).
\end{align}

The reciprocal error force, namely the difference between the
the exact and truncated reciprocal force, is
\begin{align}
  \Delta \v F^{\textrm{tr}}_{\textrm{rec}}(\v r_i)
  = & - 
  q_i
  \sum_{\vert m_\alpha\vert \geq K_\alpha/2}
  \v g(\v m) \,
  S(-\v m)\,
  e^{2\pi i\v m\cdot\v r_i} 
\end{align}
It is proved that the mean reciprocal error force is
\begin{align}\label{eqn:tmp18}
  \big\langle
  \Delta \v F^{\textrm{tr}}_{\textrm{rec}}(\v r_i)
  \big\rangle
  &= 
  -q_i\,[\v K^{\textrm{tr}}_{\textrm{rec}}\ast\rho_q\,] (\v r_i),
\end{align}
where the reciprocal convolutional kernel of the error force $\v
K^c_{\textrm{rec}}$ is defined by:
\begin{align}
  \v K^{\textrm{tr}}_{\textrm{rec}}(\v r) =
  \sum_{
      \vert m_\alpha\vert \geq K_\alpha/2}
  \v g(\v m) \,
  e^{2\pi i\v m\cdot\v r}.
\end{align}
Similarly, with the assumption of uncorrelation, it is shown that
\begin{align}
  \langle\vert\Delta\v F^{\textrm{tr}}_{\textrm{rec}}(\v r_i)\vert^2\rangle
  = &\,
  q_i^2\,[(\v K^{\textrm{tr}}_{\textrm{rec}})^2\ast\rho_{q^2}] (\v r_i) + 
  q_i^2\,[\v K^{\textrm{tr}}_{\textrm{rec}}\ast\rho_{q}]^2 (\v r_i).
\end{align}

@@

\subsection{The smooth particle mesh Ewald method}

The naive calculation of the reciprocal Ewald summation
\eqref{eqn:tmp6} will result in a computational cost of $\mathcal
O(N^2)$. It has been shown that with a careful tuning of parameter
$\beta$ and the reciprocal space truncation $K_\alpha$, the
computational cost is reduced to $\mathcal O(N^{3/2})$
\cite{perram1988asc}, which becomes not applicable when the system
size is larger than several hundreds of charged particles. Both the
particle mesh Ewald (PME) \cite{darden1993pme} and the smooth particle
mesh Ewald (SPME) \cite{essmann1995spm} methods are designed to reduce
the computational cost to $\mathcal O(N\log N)$ in nearly the same
way.  Since the SPME method is more precise \cite{deserno1998mue1} and
more popular than the PME, we will focus our discussion on the
former. All the error estimates develop by the current paper can be
easily extended to PME method.  The principle idea of the SPME method
is first to calculate the term $e^{2\pi i\v m\cdot\v r}$ on a uniform
lattice in the real space.  Therefore, all Fourier transforms can be
accerlerated by the fast Fourier transform (FFT), that is why the
computational cost of SPME is reduced to $\mathcal O(N\log N)$. Then,
for any particle position $\v r_i$, the value of $e^{2\pi i\v m\cdot\v
  r_i}$ is interpolated by the known values on the neighboring lattice
points.  The PME method uses Lagrangian interpolation, while the SPME
uses B-spline interpolation that is more precise, and has higher order
of smoothness.

The SPME method provides two possibilities of calculating the
reciprocal force. The first one is to differentiate (with respect to
the particle position) the reciprocal term of the truncated Ewald
energy and then approximate the derived force \eqref{eqn:tmp8} by the
B-spline interplation. This way is called
\emph{ik--differentiation}. The alternative way notices the high-order
smoothness of the B-spline interpolation, and derive the force by
differentiating the B-spline approximated reciprocal Ewald energy
(also truncated). This way was proposed by the original SPME paper
\cite{essmann1995spm}, and is called the \emph{analytical
  differentiation}.  It has been shown that using the same parameters,
the ik-differentiation is more precise than the analytical
differentiation, but it requres twice more FFTs, which is a bottleneck
for the communication in parallel
implementations~\cite{wang2010optimizing}. Since the SPME method is
not the point of the present paper, we refer the readers to, for
example, Ref. \cite{essmann1995spm, deserno1998mue1,
  wang2010optimizing} for details of this method.

\subsection{Error estimate of the ik-differentiation}

From the precise calculation of the reciprocal summation to the SPME
fast algorithm, the electrostatic interaction is approximated by two
steps.  The first step is the truncation of the reciprocal summation,
i.e. Eqn. \eqref{eqn:tmp6}. The second step is the approximation of
$e^{2\pi i\v m\cdot\v r}$ by B-spline
interpolation. Ref. \cite{wang2010optimizing} has pointed out that the
error due to the first step is negligible comparing with the second
step, at least for the parameters of practical interest. Numerical
examples supporting this argument in inhomogeneous systems are also
given in Section \ref{sec:tmp3}.  Therefore, to estimate the force
error, it is enough to compare the SPME force with the
\emph{truncated} Ewald force \eqref{eqn:tmp8}.

As mentioned above, the SPME error mainly oringinated from approximate
the term $e^{2\pi i\v m\cdot\v r}$ by the B-spline
interpolation. Denote the error introduced by this approximation by
$A(\v m, \v r)\,e^{2\pi i\v m\cdot \v r}$, we are fortunate to have
the analytical expression of $A(\v m, \v r)$:
\begin{align}
  A(\v m, \v r)
  =
  \sum_{\alpha=1}^3\sum_{l\neq 0}
  \frac
  {\big(\dfrac{2\pi m_\alpha}{K_\alpha} + 2\pi l\big)^{-n}}
  {\sum\limits_{l=-\infty}^{+\infty}
    \big(\dfrac{2\pi m_\alpha}{K_\alpha} + 2\pi l\big)^{-n}}
  \bigg(
  e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r} - 1
  \bigg).
\end{align}
The error force of the ik-differentiation is therfore
\begin{align}\nonumber
  \Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r_i)
  =\;&
  q_i
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \v g(\v m)\,
  A(\v m, \v r_i)\,
  e^{2\pi i\v m\cdot \v r_i}
  \sum_{j\neq i}q_j\,e^{-2\pi i\v m\cdot\v r_j} \\\nonumber
  & +\,
  q_i
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \v g(\v m)\,
  e^{2\pi i\v m\cdot \v r_i}
  \sum_{j\neq i}q_j\,A(\v m, -\v r_j)\,e^{-2\pi i\v m\cdot\v r_j}\\\nonumber
  & +\,
  q_i
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \v g(\v m)\,
  \big[
  A(\v m, \v r_i) +
  A(\v m, -\v r_i)
  \big].
\end{align}
The last ``self-interacting term'' is zero because of the symmetricity
of $\v g(\v m)$ and $A(\v m, \v r)$. 
By denoting
\begin{align}
  \v G_{\alpha,l}(\v r) =
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \v g(\v m)\,
  \frac
  {\big(\dfrac{2\pi m_\alpha}{K_\alpha} + 2\pi l\big)^{-n}}
  {\sum\limits_{l=-\infty}^{+\infty}
    \big(\dfrac{2\pi m_\alpha}{K_\alpha} + 2\pi l\big)^{-n}}
  e^{2\pi i\v m\cdot\v r}
\end{align}
And if the high wave number terms are neglected, it is proved that the
mean error force is
\begin{align}
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r_i)
  \big\rangle
  & \approx
  - 2 q_i
  \bigg[
  \big(
  \sum_{\alpha}\sum_{l\neq 0}\v G_{\alpha,l}
  \big)
  \ast\rho_q\,
  \bigg] (\v r_i)
\end{align}
Also neglect the high wave number terms, the mean squared error force is 
\begin{align}\nonumber
  \big\langle
  \vert\Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r_i)\vert^2
  \big\rangle
  \approx&\, 
  2q_i^2
  \bigg[\,
  \big(
  \sum_{\alpha} \sum_{l\neq 0}
  \vert \v G_{\alpha,l}\vert^2
  \big)
  \ast \rho_{q^2}
  \,\bigg] (\v r_i)
  +
  4q_i^2\,
  \bigg[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)^2
  \ast \rho_{q^2}\,
  \bigg] (\v r_i) \\
  &\,+
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r_i)
  \big\rangle  ^2
\end{align}

\subsection{Error estimate of the analytical differetiation}

The error force of the analytical differetiation is 

@@

  \Delta\v F_{\textrm{rec}}^{\textrm{ana}}(\v r_i)

@@

  \Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r_i)\\\nonumber

@@

  \langle \Delta\v F_{\textrm{rec}}^{\textrm{ana}}(\v r_i)\rangle

@@

  \langle \Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r_i)\rangle\\\nonumber

@@

  \langle \Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r_i)\rangle,

@@

  \vert \Delta\v F_{\textrm{rec}}^{\textrm{ana}}(\v r_i)\vert^2

@@

  \big\langle \Delta\v F_{\textrm{rec}}^{\textrm{ana}}(\v r_i)\big\rangle ^2

@@

  \big\langle \Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r_i)\big\rangle ^2 \\\nonumber

@@

  \big\langle \Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r_i)\big\rangle ^2\\

@@

\subsection{The error estimate of the total force}
